Structured Versus Decorated Cospans

### 2022-03-09 ResponseToReviews.tex

\item Beck-Chevalley condition, after Theorem A.2: Using $X_a$ instead of $C_a$ for the fibers on the diagram would be more consistent with previous 
notation.

{\bf Done.} {\john $\checkmark$} {\chris $\checkmark$}

\item COVID paragraph: I appreciate the effort into explaining the applied aspect. However, claims about the software (it is “easy”, we “easily”  modify, we “better understand”) sound too vague (is it computationally better? do we assume that the final user understands structured cospans? are we comparing  it to other Petri net editors?). Perhaps this paragraph could be fused as a comment to the previous one, which is already a very good justification  for the use of Petri nets with rates.

{\bf There is now a paper detailing how this software works, so according to this suggestion we have changed our paper so it simply refers to that other paper.}   {\john But we still need to add an arXiv link in reference 1!  It should be on the arXiv soon.}

@@

{\chris Great! The 
hyperref link ??? opens up notes by Kenny for me, I don't know how or why. I added ``according to this suggestion'' above to reflect that we indeed 
fused the paragraph as they suggested.}

{\john It's weird that the line ??? opens up anything for you.  It'll go away when we get the arXiv number.}

@@

{\bf We have added a word about Markov processes in Section 6.2, and with this, all the applications of structured or decorated cospans that we know 
of are discussed here.} {\john $\checkmark$} {\chris $\checkmark$}

\vskip 1em

{\chris Second review starts here}
\item Given the title and abstract of this paper “Structured vs Decorated Cospans” I would expect this paper to provide a judgment that structured  cospans are strictly better than decorated cospans in all circumstances and a justification for that claim, or a detailed analysis of the pros and cons of 
structured vs decorated cospans, some case studies in applications that show how the difference between these double categories is felt in the application domain, and a some description of the tradeoffs.

{\bf In fact our goal, as explained in the abstract and introduction, was to prove that the structured and decorated cospan approaches are equivalent under certain conditions, and to illustrate this result with a wide variety of examples.}

@@

{\kenny I think the reader should come to their own conclusion about which framework they should use if they are trying to decide between the two, assuming both are applicable. I wouldn't say that one is better than the other. Personally, I would rather try structured cospans first, because I would rather think about a left adjoint than a symmetric lax monoidal pseudofunctor, but other people might feel differently.}

{\john I agree with you, Kenny.  I think this reviewer saw the word ``vs'' and thought we were going to let structured and decorated cospans fight it out and announce who won, like at a boxing match.   Instead of arguing that this is not a good idea for a paper, I decided to just point out that our abstract says what our paper does, and it's something else.   (I also agree with you that structured cospans are simpler when you can use them: that's why we invented them!)}

{\john If you like the response in boldface above, please add a checkmark here.  Otherwise try to
improve it, or something.}

{\john $\checkmark$}

@@

\item Given that Petri nets with structured cospans works, and Reaction Networks work, but open dynamical systems do not, I would like some more 
explanation of the differences here. The examples of graph, labeled 
graph, Petri net, reaction net are all examples of attributed C-Sets [1], the structured cospan construction in provided for arbitrary attributed 
C-Sets in Catlab.CategoricalAlgebra [2]. The open dynamical systems are not attributed C-Sets.(...)The examples in section 6 would be much stronger 
if the others included a theorem that all (or finitary) categories of presheaves satisfy the 
conditions of theorem 4.1 or a counter example (...) Hopefully it is easy to check and provide some 
commentary 
on it in this paper.

{\john Actually it's really easy to prove that presheaf categories obey the conditions of Theorem 4.1:
given any functor $f : \mathsf{C} \to \mathsf{D}$ the left adjoint functor between presheaf categories $f_! : \widehat{\mathsf{C}} \to \widehat{\mathsf{D}}$ obeys the conditions of Theorems 4.1 and 4.2.   So I think we can easily add this as a corollary, maybe right after Theorem 4.2, to get a pile of examples.   Obvious to us, but not to everyone.  Should I do it?}

@@

\item If we take the categorical systems theory perspective advanced by section 6, and the body of the authors’ recent work on functorial semantics 
for complex 
systems, then we should ask why the decorated and structured cospan double categories are isomorphic for reaction networks, but their semantics as 
dynamical systems are a decorated cospan category that is not equivalent to any structured cospan category. Is there a different construction of 
semantics for reaction networks that uses structured cospans?

{\bf We do not know a semantics for dynamical systems that uses structured cospans; to find one would
require getting around the result at the end of Section 6.}

@@

[Several points in the text are written in a way style that can be interpreted as the dismissive style common to many discussions of Category 
Theory. I 
would recommend rewriting at least the following points that should be made in a more collegial tone.]

\item pg. 2: “But this hope is doomed for reasons painstakingly explained in [1, Section 5] … despite all the tricks one might try.” This paragraph 
provides 
no insight into the reasons that those tricks fail. Referring them to tricks is dismissive. The citation is to a paper in TAC by the current authors.

{\chris Perhaps merge this with point 16 and possibly others, and after we do something, write a common response!}

{\john He thought we were being dismissive because he didn't know that the ``tricks'' were all things I'd done myself!  If he looked at [1, Section 5] he'd see it was me coming up with a bunch of tricks to try to save the decorated cospan category of open graphs, and then Kenny and me explaining why this tricks don't work.  So maybe I should just delete ``despite all the tricks one might try''.  Sound good?}

@@

\item pg. 3: “Many of the flawed applications of decorated cospans have been fixed using structured cospans.” This dismisses the work of other 
scholars 
using the decorated cospan approach as flawed. The alleged flaw should be made more precise. What about the decorated cospan approach is inferior to 
the structured cospan approach? The goal of this paper should be to illustrate to an applied category theory audience the differences between 
decorated and structured cospans. Referring back to previous work of the authors to make that point does not provide value.

{\john The "other scholars" are mainly me and Brendan, and the flaws are that some claims are false! I think we describe the flaws in Section 6; if so 
we should point to that.}

{\john I feel like just leaving out this part in our reply: ``The goal of this paper should be to illustrate to an applied category theory audience the differences between 
decorated and structured cospans. Referring back to previous work of the authors to make that point does not provide value.''}

@@

\item pg. 27: To show that U does not have a left adjoint, we use the following well-known result. For readers of the TAC paper previously cited, 
that might 
be a well known result that does not require a citation, but my understanding of the Compositionality audience assumed that the ACT work was targeted 
readers that are more familiar with the applications and thus cannot be assumed to know such results.

{\bf We have added a citation to Emily Riehl's book ``Category Theory in Context'' which includes said result, above Lemma 6.1.} {\chris $\checkmark$}

{\john I have also changed ``well-known'' to ``known'', which is more polite to people who don't know it.} {\john $\checkmark$}

@@

{\kenny I think they are correct in that we don't use this extra assumption.}

\item p. 21, definition of F: Set $\to$ Cat

The use f, g for vertex/edge maps is opposite of the convention used on the previous page. I suggest that the same convention be used for both the 
decorated and structured approach to constructing open graphs.

{\bf At the beginning of Section 6.1, we now use $f$ as a map between vertices and $g$ as a map between edges.}

{\chris $\checkmark$} {\john $\checkmark$}

\iffalse
{\kenny I \emph{think} they are complaining that we use $f$ as a map of vertices and $g$ as a map of edges when we're talking about the functor $F \colon \mathsf{Set} \to \mathsf{Cat}$ on page 21, but we use $g$ as a map of vertices and $f$ as a map of edges in the two squares at the start of Section 6.1 on the previous page?}

@@

It seems that the left adjoint needed for structured cospans fails to exist because the chosen category of dynamical systems does not have enough 
morphisms. Relatedly, this is the only example where the pseudofunctor D: (FinSet,+) $\to$ (Cat, $\times$) comes from a simple functor D: (FinSet,+) 
$\to$ (Set, $\times$) by viewing sets as discrete categories. In the other examples, this would count as using decorated cospans in the old, flawed 
way. 
All of this suggests the question: is there a way of constructing a category of dynamical systems that has more morphisms, so that a connection with 
structured cospans can be established?

\item p. 30, following Thm A.2

If I am not mistaken, “fibration”, “opfibration,” “monoidal (op/bi)fibration” are all defined, but the definition of “bifibration” is omitted. For 
completeness, it should be included.

{\bf We have done this}

{\chris $\checkmark$} {\john $\checkmark$}

@@

{\chris I would be happy to add this to conventions, but I cannot seem to find gf anywhere in the paper. Anyone else?}

{\color{blue}{Kenny: They might not mean $gf$ specifically, but there are many instances where we don't use a $\circ$ for functional composition, e.g. 
when we're pushing forward a graph structure over a map $f$ of vertices towards the top of page 21. Do we want to use a $\circ$ or not?}}

{\chris Ah you are right. I don't mind, we could specify that for function composition we sometimes use concatenation!}

{\john There were a couple places where I put in $\circ$ --- that's what the referee is really complaining
about here.  Almost always we don't.  I could remove the $\circ$s.  There was a reason I put them in: there are some places where writing a composite using plain old concatenation looks weird to me.   But I don't think it's that big a deal.}

@@

\item p3: “Using the Grothendieck construction” - maybe worth mentioning this is the “covariant” version which is different to the most common 
convention 
(at least I think so, but the literature seems to be very split over this)

{\chris Not sure that `covariant' would make this clearer...both are very well the Grothendieck construction I would think, but I don't mind that 
much.}

\item Theorem 2.1: Maybe worth saying explicitly that s is an object of the category F(m) rather than using $\in$, since this is exactly the 
distinction to 
Fong's original definition.

{\bf  We have added a remark about this in the definition.}   \john{$\checkmark$}

@@

{\chris Maybe just specify in the statement of 2.1, definitely keep $\in$ sign in all diagrams below!}

{\john It's hard to tweak the statement, since we're really trying to introduce the notation of decorations in maps of structured cospans here!    So I've added something about this \emph{after} the statement of Theorem 2.1.  See if you like it:

\begin{quote} 
Note that the decoration $s \in F(m)$ is now an object in the category $F(m)$, not an element of a set as in Fong's original approach. 
\end{quote}

}

### 2022-03-09 cospans_new.tex
changes to cospans_new and ResponsetoReviewers

Vertical composition in $F\lCsp$ is done using composition in $\A$.  The horizontal composite of $F$-decorated cospans

@@

	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw [style=inarrow, bend left =10, looseness=1.00] (S) to (tau1);
		\draw [style=inarrow, bend left=15, looseness=1.00] (I) to (tau1);
		\draw [style=inarrow, bend left=25, looseness=1.00] (tau1) to (I);
		\draw [style=inarrow, bend left=40, looseness=1.00] (tau1) to (I);
	       \draw [style=inarrow, bend right=25, looseness=1.00] (I) to (tau2);
             \draw [style=inarrow, bend right=15, looseness=1.00] (tau2) to (R); 
\path[color=purple, very thick, shorten >=2pt, shorten <=2pt, ->, >=stealth] (1a) edge (1'a);
\path[color=purple, very thick, shorten >=2pt, shorten <=2pt, ->, >=stealth] (1b) edge (1'b);
\path[color=purple, very thick, shorten >=2pt, shorten <=2pt, ->, >=stealth] (2) edge (2');	
\path[color=purple, very thick, shorten >=2pt, shorten <=2pt, ->, >=stealth] (3) edge (3');
		\draw [style=simple] (ATL.center) to (ATR.center);
		\draw [style=simple] (ATR.center) to (ABR.center);
		\draw [style=simple] (ABR.center) to (ABL.center);
		\draw [style=simple] (ABL.center) to (ATL.center);
		\draw [style=simple] (MTL.center) to (MTR.center);
		\draw [style=simple] (MTR.center) to (MBR.center);
		\draw [style=simple] (MBR.center) to (MBL.center);
		\draw [style=simple] (MBL.center) to (MTL.center);
	\end{pgfonlayer}
\end{tikzpicture}
\]
It is an open Petri net where the transitions are labeled with rate constants $r_1, r_2 > 0$.
Here is the corresponding open dynamical system:
\begin{equation}
\label{eq:openPetrir}
  \begin{array}{ccl} \displaystyle{\frac{dS(t)}{dt}} &=& -r_1 \, S(t)I(t)  + I_1(t) + I_2(t) \\ \\
\displaystyle{\frac{dI(t)}{dt}}  &=& r_1\, S(t)I(t) - r_2 \, I(t) + I_3(t)  \\  \\
\displaystyle{\frac{dR(t)}{dt}}  &=&   r_2 \, I(t)  - O_1(t).
\end{array}
\end{equation}
Here $I_1(t),I_2(t),I_3(t)$ and $O_1(t)$ are arbitrary smooth functions of time, which describe inflows and outflows at the points $i_1,i_2,i_3 \in X$ and $o_1 \in Y$.
If we drop these inflow and outflow terms, we obtain a dynamical system: an autonomous system of coupled nonlinear first-order ordinary differential equations.   In fact these equations are a famous model of infectious disease, the `SIR model', where $S(t)$, $I(t)$ and $R(t)$ describe the populations of susceptible, infected and recovered individuals, respectively.   The inflow and outflow terms allow individuals to enter or leave the population.   This in turn lets us couple the SIR model to other models, and build larger models from smaller pieces.    Indeed, open Petri nets and the mathematics of structured cospan categories have recently been used to develop a software tool for building and manipulating epidemiological models \cite{BFMLP}.

@@

We now show that the decorated cospan double category $D \lCsp$ of open dynamical systems is not isomorphic to a structured cospan double category via \cref{thm:equiv}. Recall that in that theorem we start with the data required to build a decorated cospan category, namely a symmetric lax monoidal pseudofunctor $F \maps (\A,+) \to (\Cat,\times)$, and show that if the resulting pseudofunctor $F \maps \A \to \SMC$ factors through $\Rex$, then the opfibration $U \maps \X = \inta F \to \A$ has a left adjoint $L \maps \A \to \X$. We then obtain an isomorphism between decorated and structured cospan double categories, $F \lCsp \cong {}_L \lCsp(\X)$. We now show that in the case at hand, where $F = D$, the opfibration $U$ does \emph{not} have a left adjoint. Thus, the conditions of \cref{thm:equiv} do not hold in this case: $F$ does not factor through $\Rex$.

@@

To show that $U$ does not have a left adjoint, we use the following well-known result \cite[Lemma 4.6.1]{Riehl}:
\begin{lem} \label{lem:initial}
A functor $U \maps \A \to \X$ admits a left adjoint if and only if for every $x \in \X$, the comma category $x \downarrow U$ has an initial object.
\end{lem}
Because the empty set is initial in $\Fin\Set$, the comma category 
$\emptyset \downarrow U$ is just $\inta D$.  This contains an object $(\emptyset, v_\emptyset)$, where $v_\emptyset$ is the only possible vector field on $\R^\emptyset$, namely, the zero vector field.   The only object in $\inta D$ with any morphisms to $(\emptyset, v_\emptyset)$ is $(\emptyset, v_\emptyset)$ itself, so no other object can be initial.  However $(\emptyset, v_\emptyset)$ is not initial either, because it has no morphisms to an object $(S,v)$ unless $v$ is the zero vector field on $\R^S$.  Thus by \cref{lem:initial}, $U$ does not have a left adjoint.

>>>>

Note that the decoration $s \in F(m)$ is now an object in the category $F(m)$, not an element of a set as in Fong's original approach.  Vertical composition in $F\lCsp$ is done using composition in $\A$.  The horizontal composite of $F$-decorated cospans

@@

	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw [style=inarrow, bend left =10, looseness=1.00] (S) to (tau1);
		\draw [style=inarrow, bend left=15, looseness=1.00] (I) to (tau1);
		\draw [style=inarrow, bend left=25, looseness=1.00] (tau1) to (I);
		\draw [style=inarrow, bend left=40, looseness=1.00] (tau1) to (I);
	       \draw [style=inarrow, bend right=25, looseness=1.00] (I) to (tau2);
             \draw [style=inarrow, bend right=15, looseness=1.00] (tau2) to (R); 
\path[color=purple, very thick, shorten >=2pt, shorten <=2pt, ->, >=stealth] (1a) edge (1'a);
\path[color=purple, very thick, shorten >=2pt, shorten <=2pt, ->, >=stealth] (1b) edge (1'b);
\path[color=purple, very thick, shorten >=2pt, shorten <=2pt, ->, >=stealth] (2) edge (2');	
\path[color=purple, very thick, shorten >=2pt, shorten <=2pt, ->, >=stealth] (3) edge (3');
		\draw [style=simple] (ATL.center) to (ATR.center);
		\draw [style=simple] (ATR.center) to (ABR.center);
		\draw [style=simple] (ABR.center) to (ABL.center);
		\draw [style=simple] (ABL.center) to (ATL.center);
		\draw [style=simple] (MTL.center) to (MTR.center);
		\draw [style=simple] (MTR.center) to (MBR.center);
		\draw [style=simple] (MBR.center) to (MBL.center);
		\draw [style=simple] (MBL.center) to (MTL.center);
	\end{pgfonlayer}
\end{tikzpicture}
\]
It is an open Petri net where the transitions are labeled with rate constants $r_1, r_2 > 0$.
Here is the corresponding open dynamical system:
\begin{equation}
\label{eq:openPetrir}
  \begin{array}{ccl} \displaystyle{\frac{dS(t)}{dt}} &=& -r_1 \, S(t)I(t)  + I_1(t) + I_2(t) \\ \\
\displaystyle{\frac{dI(t)}{dt}}  &=& r_1\, S(t)I(t) - r_2 \, I(t) + I_3(t)  \\  \\
\displaystyle{\frac{dR(t)}{dt}}  &=&   r_2 \, I(t)  - O_1(t).
\end{array}
\end{equation}
Here $I_1(t),I_2(t),I_3(t)$ and $O_1(t)$ are arbitrary smooth functions of time, which describe inflows and outflows at the points $i_1,i_2,i_3 \in X$ and $o_1 \in Y$.
If we drop these inflow and outflow terms, we obtain a dynamical system: an autonomous system of coupled nonlinear first-order ordinary differential equations.   In fact these equations are a famous model of infectious disease, the `SIR model', where $S(t)$, $I(t)$ and $R(t)$ describe the populations of susceptible, infected and recovered individuals, respectively.   The inflow and outflow terms allow individuals to enter or leave the population.   This in turn lets us couple the SIR model to other models, and build larger models from smaller pieces.    Indeed, a group of researchers has recently used open Petri nets and the mathematics of structured cospans in their software tool for building and manipulating epidemiological models \cite{BFMLP}.

@@

We now show that the decorated cospan double category $D \lCsp$ of open dynamical systems is not isomorphic to a structured cospan double category via \cref{thm:equiv}. Recall that in that theorem we start with the data required to build a decorated cospan category, namely a symmetric lax monoidal pseudofunctor $F \maps (\A,+) \to (\Cat,\times)$, and show that if the resulting pseudofunctor $F \maps \A \to \SMC$ factors through $\Rex$, then the opfibration $U \maps \X = \inta F \to \A$ has a left adjoint $L \maps \A \to \X$. We then obtain an isomorphism between decorated and structured cospan double categories, $F \lCsp \cong {}_L \lCsp(\X)$. We now show that in the case at hand, where $F = D$ is the functor sending each finite set $S$ to the set of dynamical systems on $\R^S$, the opfibration $U$ does \emph{not} have a left adjoint. Thus, the conditions of \cref{thm:equiv} cannot hold in this case: $F$ does not factor through $\Rex$.

@@

To show that $U$ does not have a left adjoint, we use a known result \cite[Lemma 4.6.1]{Riehl}:
\begin{lem} \label{lem:initial}
A functor $U \maps \A \to \X$ admits a left adjoint if and only if for every $x \in \X$, the comma category $x \downarrow U$ has an initial object.
\end{lem}
Because the empty set is initial in $\Fin\Set$, the comma category 
$\emptyset \downarrow U$ is just $\inta D$.  This contains an object $(\emptyset, v_\emptyset)$, where $v_\emptyset$ is the only possible vector field on $\R^\emptyset$, namely, the zero vector field.   The only object in $\inta D$ with any morphisms to $(\emptyset, v_\emptyset)$ is $(\emptyset, v_\emptyset)$ itself, so no other object can be initial.  However $(\emptyset, v_\emptyset)$ is not initial either, because it has no morphisms to an object $(S,v)$ unless $v$ is the zero vector field on $\R^S$.  Thus by \cref{lem:initial}, $U$ does not have a left adjoint.
